Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
edits in lightning talks

\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew D. Jones
and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces to
Improve Productivity of Legacy Tools}}
Abani Patra presented an example how the community using Titan2d, a Geoflow
Simulation Software, increased the productivity of their tools by improving
both code and data layout. Main obstacles in this change were the non-existence
of a common version control system for the source code, coupled with multiple
versions of the same code base, the fixed format of input files, that many
input values used to be set as compilation flags, and that the internal data
layout was not suitable for modern technologies (vectorization, accelerators).
The approach of the Titan2d developers included reinforcing the code structure
using a multiple laters of Python and C++, and a redesign of the data layout to
be better suitable for modern CPUs and accelerators.

\item \textbf{Abigail Cabunoc Mayes, Bill Mills, Arliss Collins and Kaitlin
Thaney: \textit{Collaborative Software Development as Sustainable Software: Lessons
from Open Source}}
Abigail Cabunoc Mayes combines two properties of open source that, together,
form a suitable habitat for sustainable software. The first of the two
properties, being public, does not only mean public code. It also includes
public discussions, a public process of including contributions, and an open
license. The second property, participatory, stresses the importance of
reaching out to the community and helping potential new members by providing
better documenation and learning experiences, like code review and examples of
good first bug reports. Together, Abigail concludes, these two properties not
only lead to higher quality, reusability and ease of understanding, but
eventually, to sustainablity.

\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
through Best Practices in Open Source Software: Computational Infrastructure
for Geodynamics}}

\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
Smith and Laura Soito: \textit{Software Attribution for Geoscience Applications in the
Computational Infrastructure for Geodynamics}}
Lorraine Hwang presented experiences with CIG: the computational infrastructure
for Geodynamics, a community software with a worldwide user base. As others,
their main goals include high usability, sustainability, and reproducibility.
In order to achieve this, communication channels have been developed, like
mailing lists, wikis, workshops, hackathons, tutorials, and webinars. In order
to contribute to the infrastructure, codes have to adhere to specified minimum
standards, but ideals are also given to encourage reaching the aforementioned
goals. This includes, e.g., the use of version control, certain coding styles,
the presence and nature of code tests and documentation and certain user
workflows. In addition, various ways to cite efforts within the framework are
provided, including science papers, code papers, user manuals, and the website.
In an effort creating statistics of the resulting citations, $80\%$ of papers
that use CIG codes mention the code name, and about the same number includes a
citation. Only about $20\%$ acknowledge CIG. Within the same sample of papers,
about one fifth of use an URL to cite codes (including non-CIG codes), and only
about one eigth specifies the version used, showing that compared to other
codes, CIG seems to be much better cited. This, in part, is attributed to ideal
citation forms communicated to the community, and runtime tools that help
collecting a relevant citation list automatically.

\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
Douglas Thain and Charles Vardeman: \textit{Data And Software Preservation for Open
Science (DASPOS)}}

@@

\item \textbf{Jeffrey Carver: \textit{On the Need for Software Engineering Support for
Sustainable Scientic Software}}
Jeff argued that for scientific software to be truly sustainable, there is a need
for developers to use appropriate software engineering practices. His experience
interacting with scientific teams indicates that choosing and tailoring these
practices is not a trivial exercise. There is a general culture clash between
software engineering and science that hinders our ability to communicate and
choose appropriate methods. In addition, many experience scientific software
developers appear to be unaware of software engineering practices that may be
beneficial to them. The most appropriate software engineering practices are
those that are lightweight, properly tailored, and focus on the key software
development problems faced by scientists. In order to increase the use of
software engineering in science, we need more documented success stories. These
successes need to be socialized within the scientific community through
workshops like the Software Engineering for Science workshop
series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
and the new Software Engineering track in
\textit{Computing in Science and Engineering} magazine.

>>>>

\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew D. Jones
and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces to
Improve Productivity of Legacy Tools}}
Abani Patra presented an example how the community using Titan2D, a Geoflow
Simulation Software, increased the productivity of their tools by improving
both code and data layout. The main obstacles in this change were the non-existence
of a common version control system for the source code, coupled with multiple
versions of the same code base, the fixed format of input files, that many
input values used to be set as compilation flags, and that the internal data
layout was not suitable for modern technologies (vectorization, accelerators).
The approach of the Titan2D developers included reinforcing the code structure
using a multiple laters of Python and C++, and a redesign of the data layout to
be better suitable for modern CPUs and accelerators.

\item \textbf{Abigail Cabunoc Mayes, Bill Mills, Arliss Collins and Kaitlin
Thaney: \textit{Collaborative Software Development as Sustainable Software: Lessons
from Open Source}}
Abigail Cabunoc Mayes combines two properties of open source that, together,
create a suitable habitat for sustainable software. The first of the two
properties, public, does not only mean public code. It also includes
public discussions, a public process of including contributions, and an open
license. The second property, participatory, stresses the importance of
reaching out to the community and helping potential new members by providing
better documentation and learning experiences, like code review and examples of
good first bug reports. Together, Abigail concluded, these two properties not
only lead to higher quality, reusability and ease of understanding, but
eventually, to sustainability.

\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
through Best Practices in Open Source Software: Computational Infrastructure
for Geodynamics}}
Lorraine Hwang presented experiences with CIG: the Computational Infrastructure
for Geodynamics, a community software with a worldwide user base. Like others,
their main goals include high usability, sustainability, and reproducibility.
In order to achieve this, communication channels have been developed, such as
mailing lists, wikis, workshops, hackathons, tutorials, and webinars. In order
to contribute to the infrastructure, codes have to adhere to specified minimum
standards, but ideals are also given to encourage reaching the aforementioned
goals. This includes, e.g., the use of version control, certain coding styles,
the presence and nature of code tests and documentation and certain user
workflows.

\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
Smith and Laura Soito: \textit{Software Attribution for Geoscience Applications in the
Computational Infrastructure for Geodynamics}}
In a second talk, Lorraine Hwang mentioned various ways to cite
efforts within the framework, including science papers, code papers, user
manuals, and the website. An analysis of the resulting
citations showed that $80\%$ of papers that use CIG codes mention the code name, and about
the same number includes a citation. Only about $20\%$ acknowledge CIG. Within
the same sample of papers, about one fifth use an URL to cite codes
(including non-CIG codes), and only about one eighth specify the version used,
showing that compared to other codes, CIG seems to be much better cited. This,
in part, is attributed to suggested citation forms communicated to the community,
and runtime tools that help collecting a relevant citation list automatically.

\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
Douglas Thain and Charles Vardeman: \textit{Data And Software Preservation for Open
Science (DASPOS)}}
DASPOS is an NSF-funded multi-disciplinary effort, located at Notre
Dame and Chicago, that links the HEP effort to Biology, Astrophysics, Digital
Curation, and other disciplines. It includes physicists, digital librarians, as
well as computer scientists, and aims to achieve some commonality across
disciplines. Examples are meta-data descriptions of archived data,
computational descriptions, descriptions of how data was processed, but also
includes questions such as whether computation replication can be automated, and
what the impact of access policies on the preservation infrastructure is. One
of the products of this effort is a suite of tools that deals with this
preservation, and the questions was posed whether that software itself is
sustainable. Points that were brought up included the need for a user community
depending on a given software, and the need to provide added value for its
users. An important method to achieve this was to work with the user
community from the start, and to budget that way. For the specific example of
the preservation software, this means that besides adding value to the
community, it needs to be transparent to their workflow, i.e., not requiring
additional effort to preserve ``research objects''.

@@

\item \textbf{Jeffrey Carver: \textit{On the Need for Software Engineering Support for
Sustainable Scientific Software}}
Jeff argued that for scientific software to be truly sustainable, there is a need
for developers to use appropriate software engineering practices. His experience
interacting with scientific teams indicates that choosing and tailoring these
practices is not a trivial exercise. There is a general culture clash between
software engineering and science that hinders our ability to communicate and
choose appropriate methods. In addition, many experience scientific software
developers appear to be unaware of software engineering practices that may be
beneficial to them. The most appropriate software engineering practices are
those that are lightweight, properly tailored, and focus on the key software
development problems faced by scientists. In order to increase the use of
software engineering in science, we need more documented success stories. These
successes need to be socialized within the scientific community through
workshops like the Software Engineering for Science workshop
series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
and the new Software Engineering track in
\textit{Computing in Science and Engineering} magazine.
